---
title: "MA 615 Final Project_EDA"
author: "Jiahao Liu"
date: "2022-12-07"
output:
  pdf_document: default
  html_document: default
  word_document: default
---
```{r, warning=FALSE}
library(dplyr)
library(ggplot2)
library(fmsb)
```


I choose data from November 2021 to October 2022 to complete this EDA report.

## Calendar
```{r}
calendar_11_21 <- read.csv("C:/Users/Jiahao Liu/Desktop/calendar_attributes0.txt")
calendar_12_21 <- read.csv("C:/Users/Jiahao Liu/Desktop/calendar_attributes1.txt")
calendar_01_22 <- read.csv("C:/Users/Jiahao Liu/Desktop/calendar_attributes2.txt")
calendar_02_22 <- read.csv("C:/Users/Jiahao Liu/Desktop/calendar_attributes3.txt")
calendar_03_22_1 <- read.csv("C:/Users/Jiahao Liu/Desktop/calendar_attributes4_1.txt")
calendar_03_22_2 <- read.csv("C:/Users/Jiahao Liu/Desktop/calendar_attributes4_2.txt")
calendar_03_22_3 <- read.csv("C:/Users/Jiahao Liu/Desktop/calendar_attributes4_3.txt")
calendar_03_22 <- rbind(calendar_03_22_1, calendar_03_22_2, calendar_03_22_3)
calendar_04_22 <- read.csv("C:/Users/Jiahao Liu/Desktop/calendar_attributes5.txt")
calendar_05_22 <- read.csv("C:/Users/Jiahao Liu/Desktop/calendar_attributes6.txt")
calendar_06_22 <- read.csv("C:/Users/Jiahao Liu/Desktop/calendar_attributes7.txt")
calendar_07_22 <- read.csv("C:/Users/Jiahao Liu/Desktop/calendar_attributes8.txt")
calendar_08_22 <- read.csv("C:/Users/Jiahao Liu/Desktop/calendar_attributes9.txt")
calendar_09_22 <- read.csv("C:/Users/Jiahao Liu/Desktop/calendar_attributes10.txt")
calendar_10_22 <- read.csv("C:/Users/Jiahao Liu/Desktop/calendar_attributes11.txt")

calendar <- rbind(calendar_11_21,calendar_12_21,calendar_01_22,calendar_02_22,calendar_03_22,calendar_04_22,calendar_05_22,calendar_06_22,calendar_07_22,calendar_08_22,calendar_09_22,calendar_10_22)
```

```{r,fig.height=5,fig.width=7,fig.cap='Typicality Distribution'}
count(calendar,service_schedule_typicality)
data <- data.frame(
  category=c("Typicality-1", "Typicality-2", "Typicality-3", "Typicality-4", "Typicality-5"),
  count=c(878,4,33,472,17)
)
 
# Compute percentages
data$fraction <- data$count / sum(data$count)

# Compute the cumulative percentages (top of each rectangle)
data$ymax <- cumsum(data$fraction)

# Compute the bottom of each rectangle
data$ymin <- c(0, head(data$ymax, n=-1))

# Compute label position
data$labelPosition <- (data$ymax + data$ymin) / 2

# Compute a good label
data$label <- paste0(data$category, "\n value: ", data$count)

# Make the plot
ggplot(data, aes(ymax=ymax, ymin=ymin, xmax=4, xmin=3, fill=category)) +
  geom_rect() +
  geom_label( x=3.5, aes(y=labelPosition, label=label), size=2.5) +
  scale_fill_brewer(palette=4) +
  coord_polar(theta="y") +
  xlim(c(2, 4)) +
  theme_void() +
  theme(legend.position = "none")
  
```

```{r,fig.height=5, fig.width=7,fig.cap='Calendar Typicality'}
#unique(calendar$rating_description)
calendar_rate <- data.frame(calendar$service_schedule_typicality,calendar$rating_description)

calendar_rate_1 <- calendar_rate %>%
  filter(calendar.rating_description == 'Fall')
#count(calendar_rate_1,calendar.service_schedule_typicality)

calendar_rate_2 <- calendar_rate %>%
  filter(calendar.rating_description == 'Spring')
#count(calendar_rate_2,calendar.service_schedule_typicality)

calendar_rate_3 <- calendar_rate %>%
  filter(calendar.rating_description == 'Summer')
#count(calendar_rate_3,calendar.service_schedule_typicality)

calendar_rate_4 <- calendar_rate %>%
  filter(calendar.rating_description == 'Winter')
#count(calendar_rate_4,calendar.service_schedule_typicality)

calendar_rate_5 <- calendar_rate %>%
  filter(calendar.rating_description == 'Summer/Fall')
#count(calendar_rate_5,calendar.service_schedule_typicality)

calendar_rate_6 <- calendar_rate %>%
  filter(calendar.rating_description == 'Fall/Winter')
#count(calendar_rate_6,calendar.service_schedule_typicality)

calendar_rate_7 <- calendar_rate %>%
  filter(calendar.rating_description == 'Spring/Summer')
#count(calendar_rate_7,calendar.service_schedule_typicality)

season_rate <- as.data.frame(matrix( c(132,96,351,169,112,0,18,0,0,0,4,0,0,0,5,13,6,9,0,0,0,58,28,210,45,96,19,16,0,8,9,0,0,0,0) , ncol=5))
colnames(season_rate) <- c("Typicality-1", "Typicality-2", "Typicality-3", "Typicality-4", "Typicality-5")
rownames(season_rate) <- c("Spring", "Summer" , "Fall" , "Winter" , "Spring/Summer", "Summer/Fall", "Fall/Winter")

season_rate

library(RColorBrewer)
coul <- brewer.pal(7, "Set2")
colors_border <- coul
library(scales)
colors_in <- alpha(coul,0.4)

radarchart(season_rate,axistype=0, maxmin=F,pcol = colors_border, pfcol=colors_in, plwd=2.5, plty=1,vlcex=0.8,cglcol="grey", cglty=1)
legend(x = 'topright', legend = rownames(season_rate), bty = "n", pch=15 , col=colors_in , text.col = "grey", cex=1, pt.cex=2.5)
```
The figure shows that MBTA provide extra service in winter term, change holiday service to typical Saturday or Sunday schedule in Summer term, and in the fall term, MBTA usually change their service due to weather events or construciton.  

```{r,fig.height=5, fig.width=7, fig.cap='Week Typicality'}
week_rate <- data.frame(
  schedule = calendar$service_schedule_type, 
  typicality = calendar$service_schedule_typicality)

week_rate_1 <- week_rate %>%
  filter(schedule == 'Weekday')
#count(week_rate_1,typicality)

week_rate_2 <- week_rate %>%
  filter(schedule == 'Saturday')
#count(week_rate_2,typicality)

week_rate_3 <- week_rate %>%
  filter(schedule == 'Sunday')
#count(week_rate_3,typicality)


week_rate_new <- as.data.frame(matrix( c(366,231,281,4,0,0,5,4,24,146,162,164,17,0,0) , ncol=5))
colnames(week_rate_new) <- c("Typicality-1", "Typicality-2", "Typicality-3", "Typicality-4", "Typicality-5")
rownames(week_rate_new) <- c("Weekday", "Saturday" , "Sunday")

week_rate_new

library(RColorBrewer)
coul <- brewer.pal(3, "Set3")
colors_border <- coul
library(scales)
colors_in <- alpha(coul,0.4)

radarchart(week_rate_new,axistype=0, maxmin=F,pcol = colors_border, pfcol=colors_in, plwd=2.5, plty=1,vlcex=0.8,cglcol="grey", cglty=1)
legend(x = 'topright', legend = rownames(week_rate_new), bty = "n", pch=15 , col=colors_in , text.col = "grey", cex=1, pt.cex=2.5)
```
The figure shows that MBTA often has planned disruption due to construction on Saturday and Sunday. For the weekday service, MBTA also provide extra service.  

## Route Patterns
```{r}
route_11_21 <- read.csv("C:/Users/Jiahao Liu/Desktop/route_patterns0.txt")
route_12_21 <- read.csv("C:/Users/Jiahao Liu/Desktop/route_patterns1.txt")
route_01_22 <- read.csv("C:/Users/Jiahao Liu/Desktop/route_patterns2.txt")
route_02_22 <- read.csv("C:/Users/Jiahao Liu/Desktop/route_patterns3.txt")
route_03_22_1 <- read.csv("C:/Users/Jiahao Liu/Desktop/route_patterns4_1.txt")
route_03_22_2 <- read.csv("C:/Users/Jiahao Liu/Desktop/route_patterns4_2.txt")
route_03_22_3 <- read.csv("C:/Users/Jiahao Liu/Desktop/route_patterns4_3.txt")
route_03_22 <- rbind(route_03_22_1, route_03_22_2, route_03_22_3)
route_04_22 <- read.csv("C:/Users/Jiahao Liu/Desktop/route_patterns5.txt")
route_05_22 <- read.csv("C:/Users/Jiahao Liu/Desktop/route_patterns6.txt")
route_06_22 <- read.csv("C:/Users/Jiahao Liu/Desktop/route_patterns7.txt")
route_07_22 <- read.csv("C:/Users/Jiahao Liu/Desktop/route_patterns8.txt")
route_08_22 <- read.csv("C:/Users/Jiahao Liu/Desktop/route_patterns9.txt")
route_09_22 <- read.csv("C:/Users/Jiahao Liu/Desktop/route_patterns10.txt")
route_10_22 <- read.csv("C:/Users/Jiahao Liu/Desktop/route_patterns11.txt")

route <- rbind(route_11_21,route_12_21,route_01_22,route_02_22,route_03_22,route_04_22,route_05_22,route_06_22,route_07_22,route_08_22,route_09_22,route_10_22)
```

```{r, fig.height=5,fig.width=7, fig.cap='Typicality Distribution'}
pattern_time <- route %>% filter(
  route_pattern_time_desc == 'Early mornings only'|route_pattern_time_desc == 'Weekday evenings only'|route_pattern_time_desc == 'Saturdays only'|route_pattern_time_desc == 'School days only'
)


count(route,route_pattern_typicality)
data <- data.frame(
  category=c("Typicality-1", "Typicality-2", "Typicality-3", "Typicality-4"),
  count=c(5604,2408,2895,572 )
)
 
# Compute percentages
data$fraction <- data$count / sum(data$count)

# Compute the cumulative percentages (top of each rectangle)
data$ymax <- cumsum(data$fraction)

# Compute the bottom of each rectangle
data$ymin <- c(0, head(data$ymax, n=-1))

# Compute label position
data$labelPosition <- (data$ymax + data$ymin) / 2

# Compute a good label
data$label <- paste0(data$category, "\n value: ", data$count)

# Make the plot
ggplot(data, aes(ymax=ymax, ymin=ymin, xmax=4, xmin=3, fill=category)) +
  geom_rect() +
  geom_label( x=3.5, aes(y=labelPosition, label=label), size=5) +
  scale_fill_brewer(palette=4) +
  coord_polar(theta="y") +
  xlim(c(2, 4)) +
  theme_void() +
  theme(legend.position = "none")
```

```{r, fig,highlight=5,fig.width=7,fig.cap='Train Typicality'}
#unique(route$route_id)
route_train <- route %>% filter(
  route_id == 'Red'|route_id == 'Orange'|route_id == "Blue"|route_id == "Green-B"|route_id == 'Green-C'|route_id == 'Green-D'|route_id == 'Green-E'
)

route_train_1 <- route_train %>%
  filter(route_id == 'Red')
#count(route_train_1,route_pattern_typicality)

route_train_2 <- route_train %>%
  filter(route_id == 'Orange')
#count(route_train_2,route_pattern_typicality)

route_train_3 <- route_train %>%
  filter(route_id == 'Blue')
#count(route_train_3,route_pattern_typicality)

route_train_4 <- route_train %>%
  filter(route_id == 'Green-B')
#count(route_train_4,route_pattern_typicality)

route_train_5 <- route_train %>%
  filter(route_id == 'Green-C')
#count(route_train_5,route_pattern_typicality)

route_train_6 <- route_train %>%
  filter(route_id == 'Green-D')
#count(route_train_6,route_pattern_typicality)

route_train_7 <- route_train %>%
  filter(route_id == 'Green-E')
#count(route_train_7,route_pattern_typicality)

route_train_new <- as.data.frame(matrix( c(56,28,28,28,28,26,28,0,0,14,0,0,10,0,148,24,12,22,16,74,18) , ncol=3))
colnames(route_train_new) <- c("Typicality-1", "Typicality-3", "Typicality-4")
rownames(route_train_new) <- c("Red", "Orange" , "Blue", "Green-B", "Green-C", "Green-D", "Green-E")

route_train_new

library(RColorBrewer)
coul <- brewer.pal(7, "Set1")
colors_border <- coul
library(scales)
colors_in <- alpha(coul,0.4)

radarchart(route_train_new,axistype=0, maxmin=F,pcol = colors_border, pfcol=colors_in, plwd=2.5, plty=1,vlcex=0.8,cglcol="grey", cglty=1)
legend(x = 'topright', legend = rownames(route_train_new), bty = "n", pch=15 , col=colors_in , text.col = "grey", cex=1, pt.cex=2.5)
```
This figure shows that red line diversions from normal service, such as planned detours, bus shuttles, or snow routes. Blue line and Green-D line contains special routing which only runs a handful of times per day.
          
          